---
title: "Aim 1 - Alpha and Beta Diversity Analysis"
output: html_document
---

# AIM 1 - Alpha and Beta Diversity Analysis
# Purpose: Assess whether gut microbiome diversity differs across iron
# supplementation groups (FeSO4, MNP, None) in 12-month-old anemic infants.
# Inputs: anemia_metadata.txt, taxonomy.tsv, anemia-feature-table.txt, tree.nwk
# Outputs: alpha_diversity_main.png, supplementary_chao1_plot.png,
# supplementary_faith_pd_plot.png, bray_curtis_pcoa.png,
# unweighted_unifrac_pcoa.png

```{r libraries}
library(tidyverse)
library(phyloseq)
library(vegan)
library(ggplot2)
library(microbiome)
library(ape)
library(ggpubr)
library(picante)

```

```{r load data}

#Load metadata, taxonomy,count table, and phylogenetic tree

anemia_metadata <- read.delim("anemia_metadata.txt", row.names = 1, check.names = FALSE)
anemia_taxonomy <- read.delim("taxonomy.tsv", row.names = 1, check.names = FALSE)
anemia_counts <- read.delim("anemia-feature-table.txt", skip = 1, row.names = 1, check.names = FALSE)
tree <- read.tree("tree.nwk")

```


```{r format taxonomy and filter metadata}

# Format taxonomy table into separate taxonomic ranks

anemia_taxonomy <- anemia_taxonomy %>%
  separate(col = Taxon,
           into = c('Domain', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species'), 
           sep = ';', fill = 'right', extra = "drop") %>%
  select(-Confidence)

# Filter metadata to include only samples relevant to Aim 1: 
# 12-month-old anemic infants in MNP, FeSO4, or no supplementation groups

anemia_metadata_filtered <- anemia_metadata %>%
  filter(supplement %in% c("MNP", "FeSO4", "None"),
         age_months == 12,
         anemia == "anemic",
         !if_any(everything(), ~. == "Missing: Not collected"))

# Confirm sample distribution across supplementation groups 

table(anemia_metadata_filtered$supplement)

```

```{r match-samples}

# Match samples between metadata and count table

shared_samples <- intersect(colnames(anemia_counts), rownames(anemia_metadata_filtered))

anemia_counts <- anemia_counts[, shared_samples]
anemia_metadata <- anemia_metadata_filtered[shared_samples, ,drop = FALSE]

# Confirm sample alignment
all(colnames(anemia_counts) == rownames(anemia_metadata_filtered))

# Match taxa between count table and taxonomy table

shared_taxa <- intersect(rownames(anemia_counts), rownames(anemia_taxonomy))

anemia_counts <- anemia_counts[shared_taxa, ]
anemia_taxonomy <- anemia_taxonomy[shared_taxa, , drop = FALSE]

# Confirm taxa alignment

all(rownames(anemia_counts) == rownames(anemia_taxonomy))

```

```{r build-phyloseq}

# Build the phyloseq object used for downstream diversity analysis

OTU <- otu_table(as.matrix(anemia_counts), taxa_are_rows = TRUE)
TAX <- tax_table(as.matrix(anemia_taxonomy))
SAM <- sample_data(anemia_metadata_filtered)

ps = phyloseq(OTU, TAX, SAM, phy_tree(tree))

```

```{r rarefraction}

#Inspect sequencing depth to justify the rarefraction threshold
hist(sample_sums(ps))

#Rarefraction curve helsp visualize sampling depth across samples 
rarecurve(t(data.frame(ps@otu_table)), step = 200)

#Rarefy to an even depth of 22,500 reads per sample
#Samples below this threshold will be removed
psrare = ps %>% rarefy_even_depth(sample.size = 22500, rngseed = 421)

#Confirm equal sequencing depth after rarefraction
table(sample_sums(psrare))

```

```{r alpha diversity}

#Calculate the diversity metrics for each retained sample
alpha_plot_obj <- plot_richness(psrare, x = "supplement", measures = c("Observed", "Shannon"), color = "supplement")

#Extract the data used inside the phyloseq plot
alpha_data <- alpha_plot_obj$data

#Set supplement group order for consistent plotting
alpha_data$supplement <- factor(alpha_data$supplement, levels = c("FeSO4", "MNP", "None"))

```

```{r check normality}

# Assess whether alpha diversity metrics meet paametric assumptions
# Using Shapiro-Wilk tests and histogram inspection
alpha_normality <- alpha_data %>% 
  group_by(variable) %>%
  summarise(shapiro_p = shapiro.test(value)$p.value, 
            .groups = "drop")

alpha_normality

#Plot histograms to visually inspect metric distributions
alpha_data %>% 
    ggplot(aes(x = value)) + 
    geom_histogram(bins = 20, fill = "grey40", color = "grey40") +
    facet_wrap(~variable, scales = "free") +
    theme_classic(base_size = 14) +
    labs(x = "Alpha diversity value", y = "Count")
```

```{r alpha-kruskal} 

# Test for overall differences among FeSO4, MNP, and None groups
# for each alpha diversity metric
alpha_kw <- alpha_data %>%
  group_by(variable) %>%
  summarise(p_value = kruskal.test(value ~ supplement)$p.value, 
             .groups = "drop")

alpha_kw

```

```{r alpha-posthoc}

# Run pairwise Wilcoxon tests with BH adjsutment for post-hoc reporting
# No pairwise comparisons were significant
pairwise_alpha <- alpha_data %>% 
  group_by(variable) %>%
  group_modify(~ {
    pw <- pairwise.wilcox.test(.x$value, .x$supplement, p.adjust.method = "BH")
    
    as.data.frame(as.table(pw$p.value)) %>%
      rename(group1 = Var1, group2 = Var2, padj = Freq) %>%
      drop_na()
  }) %>% 
  ungroup()

pairwise_alpha

```

```{r alpha-plot}
# Generate final alpha diversity figure for Observed features and Shannon diversity
# No statistical annotations are because no significant differences were detected

alpha_plot <- alpha_data %>%
  ggplot(aes(x = supplement, y = value, fill = supplement)) +
  geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.85) +
  geom_jitter(width = 0.12, height = 0, size = 1.6, alpha = 0.45, color = "black") +
  facet_wrap(~variable, scales = "free_y") +
  theme_classic(base_size = 15) +
  labs(
    x = "Iron supplementation group",
    y = "Alpha diversity",
    fill = "Supplement"
  ) +
  theme(
axis.text.x = element_text(angle = 45, hjust = 1),
strip.text = element_text(face = "bold", size = 14), legend.position = "right")

alpha_plot

```

```{r make-bh-annotation-table}
# Create labels for BH-adjusted p-values
# No pairwise comparisons are significant


maxvals <- alpha_data %>%
  group_by(variable) %>%
  summarise(max_y = max(value), .groups = "drop")

annot <- pairwise_alpha %>%
  mutate(
    xmin = case_when(
      group1 == "MNP"  & group2 == "FeSO4" ~ 2,
      group1 == "None" & group2 == "FeSO4" ~ 1,
      group1 == "None" & group2 == "MNP"   ~ 2
    ),
    xmax = case_when(
      group1 == "MNP"  & group2 == "FeSO4" ~ 1,
      group1 == "None" & group2 == "FeSO4" ~ 3,
      group1 == "None" & group2 == "MNP"   ~ 3
    )
  ) %>%
  left_join(maxvals, by = "variable") %>%
  group_by(variable) %>%
  mutate(
    y = c(1.10, 1.20, 1.30) * max_y,
    label = paste0("ns, p = ", signif(padj, 2))
  ) %>%
  ungroup()

annot
```

```{r alpha-plot-bh}
# Clean publication-style alpha diversity plot
#No significant differences were detected

alpha_plot_bh <- alpha_plot +
  geom_segment(
    data = annot,
    aes(x = xmin, xend = xmax, y = y, yend = y),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  geom_segment(
    data = annot,
    aes(x = xmin, xend = xmin, y = y, yend = y - 0.05 * max_y),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  geom_segment(
    data = annot,
    aes(x = xmax, xend = xmax, y = y, yend = y - 0.05 * max_y),
    inherit.aes = FALSE,
    linewidth = 0.5
  ) +
  geom_text(
    data = annot,
    aes(x = (xmin + xmax) / 2, y = y + 0.05 * max_y, label = label),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.30)))

alpha_plot_bh

```
```{r save-plot}
# Save high-resolution figure for manuscript/assignment use
ggsave(
  filename = "aim1_alpha_observed_shannon.png",
  plot = alpha_plot_bh,
  width = 10,
  height = 6,
  dpi = 600
)
```

```{r chao1-data} 

# Calculate Chao1 alpha diversity from the rarefied phyloseq object 

chao1_plot_obj <- plot_richness(psrare,
                  x = "supplement", 
                  measures = c("Chao1"),
                  color = "supplement")

chao1_data <- chao1_plot_obj$data

# Set supplementation group order for consistent plotting
chao1_data$supplement <-     factor(chao1_data$supplement, 
  levels = c("FeSO4", "MNP", "None"))

```

```{r chao1-plot} 

# Generate the final Chao1 alpha diversity figure

p_chao1 <- chao1_data %>%
  ggplot(aes(x = supplement, y = value, fill = supplement)) + geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.85) + geom_jitter(width = 0.12, height = 0, size = 1.6, alpha = 0.45, color = "black") + theme_classic(base_size = 15) + labs(title = "Chao1 alpha diversity", x = "Iron supplementation group", y = "Chao1 richness estimate", fill = "Supplement") + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")

p_chao1
```

```{r chao1-stats}

# Assess normality within supplementation groups
chao1_normality <- chao1_data %>%
    group_by(supplement) %>%
    summarise(shapiro_p = shapiro.test(value)$p.value, .groups = "drop")

chao1_normality

# Test for overall differences across supplementation groups
chao1_kw <- kruskal.test(value ~ supplement, data = chao1_data)
chao1_kw

# Run pairwise Wilcoxon tests with BH adjustment for post-hoc reporting
chao1_pw <- pairwise.wilcox.test(chao1_data$value, chao1_data$supplement, p.adjust.method = "BH", exact = FALSE)

# No significant differences were detected across supplementation groups 

# Convert to dataframe for plotting

chao1_annot <- as.data.frame(as.table(chao1_pw$p.value)) %>% rename(group1 = Var1, group2 = Var2, padj = Freq) %>% drop_na ()

chao1_plot_labeled <- p_chao1 +
  stat_compare_means(comparisons = list( c("FeSO4", "MNP"),
c("FeSO4", "None"),
c("MNP", "None")),
method = "wilcox.test",
label = "p.format")

chao1_plot_labeled
```
```{r faith-data}

# Calculate Faith's phylogenetic diversity from the rareified phyloseq object

# Convert OTU table so taxa are columns
otu <- as(otu_table(psrare), "matrix")
if(taxa_are_rows(psrare) == TRUE) otu <- t(otu)

# Extract phlogenetic tree from phyloseq object
faith_tree <- phy_tree(psrare)

# Check ASV names match between OTU table and tree
stopifnot(all(colnames(otu) %in% faith_tree$tip.label))

# Drop unmatched tips and reorder tree tips to match OTU table
faith_tree <- drop.tip(tree, setdiff(tree$tip.label, colnames(otu)))
faith_tree <- keep.tip(tree, colnames(otu))

# Root the tree if needed
if (!is.rooted(tree)) {tree <- root(tree, outgroup = tree$tip.label[1], resolve.root = TRUE)}

# Convert abundance to presence/absence for Faith's PD calculation
otu_pa <- otu
otu_pa[otu_pa > 0] <- 1

# Calculate Faith's phylogenetic diversity
faith <- pd(otu_pa, tree, include.root = TRUE) %>% select(-SR) %>% rownames_to_column("SampleID") %>% rename(Faith_PD = PD)

```

```{r faith-merge}

# Merge Faith's PD values with sample metadata
faith_data <- faith %>%
  left_join(sample_data(psrare) %>%
              data.frame() %>%
              rownames_to_column("SampleID"),
            by = "SampleID")

# Set supplementation group order for consistent plotting
faith_data$supplement <- factor(faith_data$supplement,
      levels = c("FeSO4", "MNP", "None"))

```

```{r faith-plot}

# Generate the final Faith's PD figure
p_faith <- faith_data %>%
  ggplot(aes(x = supplement, y = Faith_PD, fill = supplement)) +
  geom_boxplot(outlier.shape = NA, width = 0.65, alpha = 0.85) +
  geom_jitter(width = 0.12, height = 0, size = 1.6, alpha = 0.45, color = "black") + theme_classic(base_size = 15) +
  labs(title = "Faith's phylogenetic diversity", x = "Iron supplementation group", y = "Faith's PD", fill = "Supplement") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")

p_faith

faith_plot_labeled <- p_faith +
  stat_compare_means(
    comparisons = list(
      c("FeSO4", "MNP"),
      c("FeSO4", "None"),
      c("MNP", "None")),
  method = "t.test",
  label = "p.format")

faith_plot_labeled

```

```{r faith-stats}

# Check normality within each supplementation group

faith_normality <- faith_data %>%
    group_by(supplement) %>%
    summarise(shapiro_p = shapiro.test(Faith_PD)$p.value, .groups = "drop")

faith_normality

# Faith's PD met parametric assumptions (Shapiro-Wilk test, p > 0.05 for all groups),
# so differences across supplementation groups were assessed using one-way ANOVA.

anova_faith <- aov(Faith_PD ~ supplement, data = faith_data)
summary(anova_faith)

faith_pw <- pairwise.t.test(faith_data$Faith_PD,
                            faith_data$supplement, 
                            p.adjust.method = "BH")
faith_pw

# No significant differences detected across supplementation groups
```

```{r save-supp-alpha-plots}

#Save supplementary figures

ggsave("aim1_alpha_chao1.png", plot = chao1_plot_labeled, width = 7, height = 5, dpi = 600)
ggsave("aim1_alpha_faith_pd.png", plot = faith_plot_labeled, width = 7, height = 5, dpi = 600)
```

```{r beta-bray-pcoa}

# Calculate Bray-Curtis distances and run PCoA
ps_bray <- phyloseq::distance(psrare, method = "bray")
ord_bray <- ordinate(psrare, method = "PCoA", distance = ps_bray)

# Extract PCoA coordinates and merge with sample metadata
bray_data <- ord_bray$vectors %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(sample_data(psrare) %>% data.frame() %>% rownames_to_column("SampleID"), by = "SampleID")

#Extract percent variance explained by the first two axes
bray_var <- ord_bray$values$Relative_eig * 100 
```

```{r beta-bray-plot}
# Plot Bray-Curtis PCoA

p_bray <- bray_data %>%
  ggplot(aes(x = Axis.1, y = Axis.2, color = supplement)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse() +
  theme_classic(base_size = 15) +
  labs(title = "Bray-Curtis beta diversity",
        x = paste0("Axis.1 (", round(bray_var[1], 1), "%)"),
        y = paste0("Axis.2 (", round(bray_var[2], 1), "%)"),
        color = "Supplement")

p_bray
```

```{r beta-bray-permanova}

#Run Permanova to test whether Bray-Curtis community composition differs across supplementation groups
meta_df <- sample_data(psrare) %>% data.frame()
bray_perm <- adonis2(ps_bray ~ supplement, data = meta_df)

bray_perm
```

```{r beta-unifrac-distance}

# Calculate Unweighted UniFrac distance matrix
# This measures between-sample community dissimilarity using
# phylogeny and presence/absence information
ps_unifrac <- phyloseq::distance(psrare, method = "unifrac", weighted = FALSE)

```

```{r beta-unifrac-pcoa}

# Run PCoA on Unweighted UniFrac distance matrix
ord_unifrac <- ordinate(psrare, method = "PCoA", distance = ps_unifrac)

#Extract PCoA coordinates and merge with metadata 
unifrac_data <- ord_unifrac$vectors %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  left_join(
    sample_data(psrare) %>%
      data.frame() %>%
      rownames_to_column("SampleID"),
    by = "SampleID")

#Extract percent variance explained by the first two axes
unifrac_var <- ord_unifrac$values$Relative_eig * 100
```

```{r beta-unifrac-plot}

#Plot Unweighted UniFrac PCoA

p_unifrac <- unifrac_data %>%
  ggplot(aes(x = Axis.1, y = Axis.2, color = supplement)) +
  geom_point(size = 3, alpha = 0.8) +
  stat_ellipse() +
  theme_classic(base_size = 15) +
  labs( title = "Unweighted UniFrac beta diversity", 
        x = paste0("PCoA1 (", round(unifrac_var[1], 1), "%)"),
        y = paste0("PCoA2 (", round(unifrac_var[2], 1), "%)"),
        color = "Supplement")

p_unifrac
```

```{r beta-unifrac-permanova}

#Run PERMANOVA to test whether Unweighted UniFrac community composition differs across supplementation groups
unifrac_perm <- adonis2(ps_unifrac ~ supplement, data = meta_df)

unifrac_perm
```

```{r save-beta-plots}

ggsave("aim1_beta_bray_pcoa.png", plot = p_bray, width = 6, height = 5, dpi = 600)
ggsave("aim1_beta_unweighted_unifrac_pcoa.png", plot = p_unifrac, width = 6, height = 5, dpi = 600)
```