---
title: "Aim4"
output: html_document
---

# 1. Loading libraries

```{r}
library(phyloseq)
library(tidyverse)
library(broom)
library(purrr)
library(ggplot2)
```

# 2. Using the results in alpha_data (contains the Shannon index from alpha diversity analysis) from AIM1_alpha_beta_diversity.Rmd, filter for only the Shannon index).

```{r}
shannon_filter = alpha_data %>%
    filter(variable == "Shannon")
```

# 3. Fitting linear regression models for ferritin and hemoglobin levels (all supplemented/non-supplemented infants are grouped together)

-   Linear model for hemoglobin (tests to see if there is a correlation between Shannon index and hemoglobin levels)

```{r}
shannon_filter$hemoglobin <- as.numeric(shannon_filter$hemoglobin) # converting hemoglobin to numeric values

# for a Spearman correlation, use method='spearman'
cor.test(shannon_filter$hemoglobin, shannon_filter$value, method="spearman", exact = FALSE)

#p-value = 0.00611 and rho = 0.3863268, so there is a significant difference between Shannon index and hemoglobin when all the supplemented/non-supplemented infants are grouped together.
```

-   Linear model for ferritin (tests to see if there is a correlation between Shannon index and ferritin levels)

```{r}
shannon_filter$ferritin <- as.numeric(shannon_filter$ferritin) # converting ferritin to numeric values

# for a Spearman correlation, use method='spearman'
cor.test(shannon_filter$ferritin, shannon_filter$value, method="spearman", exact = FALSE)

#p-value = 0.004812 and rho = 0.3963569, so there is a significant difference between Shannon index and ferritin when all the supplemented/non-supplemented infants are grouped together.

```

# 4. Fitting linear regression models for ferritin and hemoglobin levels (the supplemented/non-supplemented infants are not grouped together)

-   Linear model for hemoglobin (tests to see if the association between the Shannon index and hemoglobin levels are influenced by supplement type)

```{r}
shannon_filter$supplement = relevel(factor(shannon_filter$supplement), ref = "None")

supplement_hemoglobin_mdl <- lm(value ~ hemoglobin*supplement, data = shannon_filter)

summary(supplement_hemoglobin_mdl)

#p-values are all above 0.05, so there is no significant difference between Shannon index and hemoglobin levels when the supplementation type changes compared to the non-supplemented group. 
```

-   Linear model for ferritin (tests to see if the association between the Shannon index and ferritin levels are influenced by supplement type)

```{r}
supplement_ferritin_mdl <- lm(value ~ ferritin*supplement, data = shannon_filter)

summary(supplement_ferritin_mdl)
#p-values are all above 0.05, so there is no significant difference between Shannon index and ferritin levels when the supplementation type changes compared to the non-supplemented group. 
```

# 5. Visualizing linear regression models for hemoglobin and ferritin

-   Hemoglobin plots

```{r}
# Grouped supplements
grouped_hemoglobin_lm_plot = ggplot(shannon_filter, aes(x = hemoglobin, y = value)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(x = "Hemoglobin (g/L)", 
         y = "Shannon Index", 
         title = "Linear Regression of Hemoglobin vs. Shannon Index", 
         caption = "p-value = 0.00611, rho = 0.3863268") +
    theme_gray(base_size = 18)

grouped_hemoglobin_lm_plot

ggsave("Grouped Hemoglobin Plot.png",
       plot = grouped_hemoglobin_lm_plot, width = 10, height = 7)


# Individual supplements
supplement_hemoglobin_lm_plot = ggplot(shannon_filter, aes(x = hemoglobin, y = value)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(.~ supplement) +
    labs(x = "Hemoglobin (g/L)", y = "Shannon Index", title = "Linear Regression of Hemoglobin vs. Shannon Index based on Supplementation Type") +
    theme_gray(base_size = 18)

supplement_hemoglobin_lm_plot

ggsave("Supplement Hemoglobin Plot.png",
       plot = supplement_hemoglobin_lm_plot, width = 10, height = 5)

```

-   Ferritin plots

```{r}
# Grouped supplements
grouped_ferritin_lm_plot = ggplot(shannon_filter, aes(x = ferritin, y = value)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(x = "Ferritin (μg/L)", 
         y = "Shannon Index", 
         title = "Linear Regression of Ferritin vs. Shannon Index",
         caption = "p-value = 0.004812, rho = 0.3963569") +
    theme_gray(base_size = 18)

grouped_ferritin_lm_plot

ggsave("Grouped Ferritin Plot.png",
       plot = grouped_ferritin_lm_plot, width = 10, height = 7)


# Individual supplements
supplement_ferritin_lm_plot = ggplot(shannon_filter, aes(x = ferritin, y = value)) +
    geom_point() +
    geom_smooth(method = "lm") +
    facet_wrap(.~ supplement) +
    labs(x = "Ferritin (μg/L)", y = "Shannon Index", title = "Linear Regression of Ferritin vs. Shannon Index based on Supplementation Type") +
    theme_gray(base_size = 18)

supplement_ferritin_lm_plot

ggsave("Supplement Ferritin Plot.png",
       plot = supplement_ferritin_lm_plot, width = 10, height = 5)
```
